### 2019-12-11 ack.tex
update

To Amor, a big thank you for contributing countless cuddles and endless drool.\\I guesstimate at least 50\% of the dissertation was written with you on my lap.

Hello Petit Remi! You're only seven months old,\\but it feels like I've known you for much longer already.
Your spontaneous happiness when you see me come downstairs\\from a hard day of work immediately makes my day.
Dear Caro, I cannot imagine a better partner and mother than you.\\You make me want to be the best version of myself.
Let's get back on track collecting doorsteps, and spending time doing the things we love,\\just us three. I look forward to the rest of my life together with my strong independent woman.

>>>>

To Amor, a big thank you for contributing countless cuddles and endless drool.
I guesstimate at least 50\% of the dissertation was written with you on my lap.
Hello Petit Remi! You're only seven months old,\\but it feels like I've known you for much longer already.
Your spontaneous happiness when you see me come downstairs\\from a hard day of work immediately makes my day.
Dear Caro, I cannot imagine a better partner and mother than you.\\You make me want to be the best version of myself.
Let's get back on track collecting doorsteps,\\and doing the things we love doing, just us three.

### 2019-12-11 ch10_discussion/content.tex
update

In \textbf{casewise NI}, one regulatory network is predicted for every single-cell in the dataset. Doing so allows studying the heterogeneity (or homogeneity) between and within different cell populations. In essence, most analyses that can be run on single-cell omics (e.g. diffential expression) can also be applied on casewise regulatory networks (e.g. to predict differentially expressed interactions).
It is the most generalised form of NI of the variants discussed above, as a casewise regulatory network can be clustered to create a differential network, or trajectory inference can be applied to derive a dynamic regulatory network. Examples of casewise NI methods include SCENIC \cite{aibar_scenicsinglecellregulatory_2017}, SSN \cite{liu_personalizedcharacterizationdiseases_2016}, and LIONESS \cite{kuijjer_estimatingsamplespecificregulatory_2019}.

>>>>

In \textbf{casewise NI}, one regulatory network is predicted for every case (sample, patient, single cell) in the dataset. Such methodology allows generating hypotheses of deregulated interactions for individual patients, which could be used for drug re-purposing and personalised medicine. Furthermore, most analyses that can be run on single-cell omics (e.g. diffential expression) can also be applied on casewise regulatory networks (e.g. to predict differentially expressed interactions).
It is the most generalised form of NI of the variants discussed above, as a casewise regulatory network can be clustered to create a differential network, or trajectory inference can be applied to derive a dynamic regulatory network. Examples of casewise NI methods include SCENIC \cite{aibar_scenicsinglecellregulatory_2017}, SSN \cite{liu_personalizedcharacterizationdiseases_2016}, and LIONESS \cite{kuijjer_estimatingsamplespecificregulatory_2019}.

### 2019-12-11 ch1_introduction/content.tex
update

\subsection{Cell types}
Ever since Robert Hook first described the different structures of cells in 1665, biologists have been classifying cells by form and function.
The human body is said to contain more than 210 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous. This however, is a major underestimation of the real number of cell types. Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types \cite{strachan_humanmoleculargenetics_2011}.

@@

One common approach for understanding the functionality of a particular cell is to observe which molecules are present in the cell and to associate those set of molecules with functionality. Taking a snap shot of the protein or RNA transcript content in a particular cell, might already provide us with major insights into its functionality. However, in order to fulfil a particular task, the biochemical machinery of the cell gradually changes over time. Therefore it is highly informative to also consider the transition states between cell types and the dynamic processes involved therein.

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/selfassessment/self_assessment.pdf}
	\caption{
		\textbf{Less than half of all TI articles perform quantitative self-assessment.}
		\textbf{A:} Since 2016, the number of TI articles has been increasing rapidly. Note that TI methods with both a pre-print and a peer-reviewed article only count once in the overall tally.
		\textbf{B:} Less than 50\% of articles feature a self-assessment. Peer-reviewed articles self-assess only in 34\% of cases.
	}
	\label{fig:benchmarks_over_time}
\end{figure}

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/overview.pdf}
	\caption{Overview}
	\label{fig:overview_thesis}
\end{figure}

>>>>

\subsection{Cell types}
Ever since Robert Hook first described the different structures of cells in 1665, biologists have been classifying cells into different "cell types", by grouping them according to form and function.
The human body is said to contain more than 210 different cell types that are classified into four groups: epithelial, connective, muscle, and nervous. This however, is a major underestimation of the real number of cell types. Neurons, for instance, that are known to be extremely diverse, are estimated to reach numbers above 10,000 different types \cite{strachan_humanmoleculargenetics_2011}.

@@

One common approach for understanding the functionality of a particular cell is to observe which molecules are present in the cell and to associate those set of molecules with functionality. Taking a snapshot of the protein or RNA transcript content in a particular cell, might already provide us with major insights into its functionality. However, in order to fulfil a particular task, the biochemical machinery of the cell gradually changes over time. Therefore it is highly informative to also consider the transition states between cell types and the dynamic processes involved therein.

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/selfassessment/self_assessment_edited.pdf}
	\caption{
		\textbf{Less than half of all TI articles perform quantitative self-assessment.}
		\textbf{A:} Since 2016, the number of TI articles has been increasing rapidly. Note that TI methods with both a pre-print and a peer-reviewed article only count once in the overall tally.
		\textbf{B:} Less than 50\% of articles feature a self-assessment. Peer-reviewed articles self-assess only in 34\% of cases.
	}
	\label{fig:benchmarks_over_time}
\end{figure}

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.75\linewidth]{fig/overview_cropped.pdf}
	\caption{Overview}
	\label{fig:overview_thesis}
\end{figure}

### 2019-12-11 ch3_dynbenchmark/content.tex
update

In this study, we evaluated the accuracy, scalability, stability and usability of 45 TI methods (Figure \ref{fig:figure_1}a). We found substantial complementarity between current methods, with different sets of methods performing most optimally depending on the characteristics of the data. For method users, we created an interactive set of guidelines (available at \href{http://guidelines.dynverse.org}{guidelines.dynverse.org}), which gives context-specific recommendations for method usage. Our evaluation also highlights some challenges for current methods, and our evaluation strategy can be useful to spearhead the development of new tools that accurately infer trajectories on ever more complex use cases.

@@

Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories (Figure \ref{fig:figure_6}a,b). However, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable tool for analyzing single-cell omics datasets with complex trajectories. Foremost, new methods should focus on improving the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory could easily be identified using a dimensionality reduction method (Figure \ref{fig:figure_6}c,d). Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell omics field. Finally, new tools should be designed to scale well with the increasing number of cells and features. We found that only a handful of current methods can handle datasets with more than 10,000 cells within a reasonable time frame. To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method on the different measures proposed in this study at \href{https://benchmark.dynverse.org}{https://benchmark.dynverse.org}.

@@

Methods were excluded from the evaluation based on several criteria: (1) not freely available; (2) no code available; (3) superseded by another method; (4) requires data types other than expression; (5) no programming interface; (6) unresolved errors during wrapping; (7) too slow (requires more than 1 h on a 100 $\times$ 100 dataset); (8) does not return an ordering; and (9) requires additional user input during the algorithm (other than prior information). The discussions on why these methods were excluded can be found at \href{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}. In the end, we included 45 methods in the evaluation.

@@

To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and Singularity containers (available at \href{https://methods.dynverse.org}{https://methods.dynverse.org}). These containers are automatically built and tested using Travis continuous integration (\href{https://travis-ci.org/dynverse}{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity. For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script reads in the input data, runs the method and outputs the files required to construct a trajectory. We also created a script to generate an example dataset, which is used for automated testing.

We used the Github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata and the usability scores. About one-third of the authors responded and we improved the wrappers based on their feedback. These discussions can be viewed on Github: \href{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}{https://github.com/dynverse/dynmethods/issues?q=label:method\_discussion}

@@

\begin{itemize}
	\item \textbf{CALISTA}: We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest. The cluster probabilities of the two selected milestones were then used as milestone percentages. This was then processed as a type 1, direct, method.
	\item \textbf{DPT}: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contained the cells that were assigned to the ‘unknown’ branch) and three terminal milestones, each corresponding to a tip point. This was then processed as a type 1, direct, method.
	\item \textbf{Sincell}: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, with the default cutoff set to 25$\%$. This was then processed as a type 7, cell graph, method.
	\item \textbf{SLICE}: As discussed in the vignette of SLICE (\href{https://research.cchmc.org/pbge/slice.html}{https://research.cchmc.org/pbge/slice.html}), we ran principal curves one by one for every edge detected by SLICE. This was then processed as a type 1, direct, method.
	\item \textbf{MFA}: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as a type 4, end state probabilities, method.
	\item \textbf{URD}: We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.

@@

More information on how each method was wrapped can be found within the comments of each wrapper script, listed at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.

@@

We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\hyperref[tab:supptable_2]{3.2}). The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalized and filtered using a common pipeline, as discussed later. Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation. In the end, we included 110 datasets for this evaluation.

@@

We used the same approach for selecting the best set of methods in the guidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}), developed using the R shiny package. This app will also filter the methods, among other things, depending on the predicted running time and memory requirements, the prior information available and the preferred execution environment (using the dynmethods package or standalone).

@@

Further information on research design is available in the Nature Research Reporting Summary, available at \href{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}.

@@

The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_himglocalmetric_2015} which was adopted from the R nettools package (\href{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}). It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network. Conceptually, \textit{HIM} is a linear combination of:

@@

To calculate the $\textit{cor}_{\textit{features}}$ we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory. More specifically, we first calculated the geodesic distances for each cell to all milestones in the trajectory. Next, we trained a Random Forest regression model (implemented in the R \textit{ranger} package \cite{wright_rangerfastimplementation_2017}, \href{https://github.com/imbs-hl/ranger}{https://github.com/imbs-hl/ranger}) to predict these distances for each milestone, based on the expression of genes within each cell. We then extracted feature importances using the Mean Decrease in Impurity (importance = 'impurity' parameter of the ranger function), as illustrated in Figure \ref{fig:snote1fig_8}. The overall importance of a feature (gene) was then equal to the mean importance over all milestones. Finally, we compared the two rankings by calculating the Pearson correlation, with values between -1 and 0 clipped to 0.

@@

We generated a panel of toy datasets (using our \href{https://github.com/dynverse/dyntoy}{\textit{dyntoy}} package, \url{https://github.com/dynverse/dyntoy}) with all possible combinations of:

@@

We then perturbed the trajectories in these datasets in certain ways, and tested whether the scores follow an expected pattern. An overview of the conformity of every metric is first given in Table \ref{tab:conformity_overview}. The individual rules and metric behaviour are discussed in the Supplementary Material that can be found at \href{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}.

@@

We do however want to stress that different use cases will require a different overall score to order the methods. Such a context-dependent ranking of all methods is provided through the dynguidelines app (\href{http://guidelines.dynverse.org}{http://guidelines.dynverse.org}).

>>>>

In this study, we evaluated the accuracy, scalability, stability and usability of 45 TI methods (Figure \ref{fig:figure_1}a). We found substantial complementarity between current methods, with different sets of methods performing most optimally depending on the characteristics of the data. For method users, we created an interactive set of guidelines (available at \url{http://guidelines.dynverse.org}), which gives context-specific recommendations for method usage. Our evaluation also highlights some challenges for current methods, and our evaluation strategy can be useful to spearhead the development of new tools that accurately infer trajectories on ever more complex use cases.

@@

Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories (Figure \ref{fig:figure_6}a,b). However, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable tool for analyzing single-cell omics datasets with complex trajectories. Foremost, new methods should focus on improving the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory could easily be identified using a dimensionality reduction method (Figure \ref{fig:figure_6}c,d). Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell omics field. Finally, new tools should be designed to scale well with the increasing number of cells and features. We found that only a handful of current methods can handle datasets with more than 10,000 cells within a reasonable time frame. To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method on the different measures proposed in this study at \url{https://benchmark.dynverse.org}.

@@

Methods were excluded from the evaluation based on several criteria: (1) not freely available; (2) no code available; (3) superseded by another method; (4) requires data types other than expression; (5) no programming interface; (6) unresolved errors during wrapping; (7) too slow (requires more than 1 h on a 100 $\times$ 100 dataset); (8) does not return an ordering; and (9) requires additional user input during the algorithm (other than prior information). The discussions on why these methods were excluded can be found at \url{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}. In the end, we included 45 methods in the evaluation.

@@

To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and Singularity containers (available at \url{https://methods.dynverse.org}). These containers are automatically built and tested using Travis continuous integration (\url{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity. For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script reads in the input data, runs the method and outputs the files required to construct a trajectory. We also created a script to generate an example dataset, which is used for automated testing.

We used the Github issues system to contact the authors of each method, and asked for feedback on the wrappers, the metadata and the usability scores. About one-third of the authors responded and we improved the wrappers based on their feedback. These discussions can be viewed on Github: \url{https://github.com/dynverse/dynmethods/issues?q=label:method_discussion}

@@

\begin{itemize}
	\item \textbf{CALISTA}: We assigned the cells to the branch at which the sum of the cluster probabilities of two connected milestones was the highest. The cluster probabilities of the two selected milestones were then used as milestone percentages. This was then processed as a type 1, direct, method.
	\item \textbf{DPT}: We projected the cells onto the cluster network, consisting of a central milestone (this cluster contained the cells that were assigned to the ‘unknown’ branch) and three terminal milestones, each corresponding to a tip point. This was then processed as a type 1, direct, method.
	\item \textbf{Sincell}: To constrain the number of milestones this method creates, we merged two cell clusters iteratively until the percentage of leaf nodes was below a certain cutoff, with the default cutoff set to 25$\%$. This was then processed as a type 7, cell graph, method.
	\item \textbf{SLICE}: As discussed in the vignette of SLICE (\url{https://research.cchmc.org/pbge/slice.html}), we ran principal curves one by one for every edge detected by SLICE. This was then processed as a type 1, direct, method.
	\item \textbf{MFA}: We used the branch assignment as state probabilities, which together with the global pseudotime were processed as a type 4, end state probabilities, method.
	\item \textbf{URD}: We extracted the pseudotime of a cell within each branch using the y positions in the tree layout. This was then further processed as a type 1, direct, method.

@@

More information on how each method was wrapped can be found within the comments of each wrapper script, listed at \url{https://methods.dynverse.org}.

@@

We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\hyperref[tab:supptable_2]{3.2}). The scripts to download and process these datasets are available on our repository (\url{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalized and filtered using a common pipeline, as discussed later. Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation. In the end, we included 110 datasets for this evaluation.

@@

We used the same approach for selecting the best set of methods in the guidelines app (\url{http://guidelines.dynverse.org}), developed using the R shiny package. This app will also filter the methods, among other things, depending on the predicted running time and memory requirements, the prior information available and the preferred execution environment (using the dynmethods package or standalone).

@@

Further information on research design is available in the Nature Research Reporting Summary, available at \url{https://www.nature.com/articles/s41587-019-0071-9\#MOESM2}.

@@

The \textit{HIM} metric (Hamming-Ipsen-Mikhailov distance) \cite{jurman_himglocalmetric_2015} which was adopted from the R nettools package (\url{https://github.com/filosi/nettools}{https://github.com/filosi/nettools}). It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network. Conceptually, \textit{HIM} is a linear combination of:

@@

To calculate the $\textit{cor}_{\textit{features}}$ we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory. More specifically, we first calculated the geodesic distances for each cell to all milestones in the trajectory. Next, we trained a Random Forest regression model (implemented in the R \textit{ranger} package \cite{wright_rangerfastimplementation_2017}, \url{https://github.com/imbs-hl/ranger}) to predict these distances for each milestone, based on the expression of genes within each cell. We then extracted feature importances using the Mean Decrease in Impurity (importance = 'impurity' parameter of the ranger function), as illustrated in Figure \ref{fig:snote1fig_8}. The overall importance of a feature (gene) was then equal to the mean importance over all milestones. Finally, we compared the two rankings by calculating the Pearson correlation, with values between -1 and 0 clipped to 0.

@@

We generated a panel of toy datasets (using our \textit{dyntoy} package, \url{https://github.com/dynverse/dyntoy}) with all possible combinations of:

@@

We then perturbed the trajectories in these datasets in certain ways, and tested whether the scores follow an expected pattern. An overview of the conformity of every metric is first given in Table \ref{tab:conformity_overview}. The individual rules and metric behaviour are discussed in the Supplementary Material that can be found at \url{https://www.nature.com/articles/s41587-019-0071-9\#Sec34}.

@@

We do however want to stress that different use cases will require a different overall score to order the methods. Such a context-dependent ranking of all methods is provided through the dynguidelines app (\url{http://guidelines.dynverse.org}).
